Concurrent Programming with Harmony

### 2020-07-28 book/paper.tex
Made a pass over the first four chapters

Harmony programs are compiled to Harmony bytecode
\index{bytecode}%
(machine instructions for a virtual machine),
which in turn is executed by the Harmony virtual machine (HVM).
\index{virtual machine}%
\index{Harmony Virtual Machine}%
\index{HVM}%
To understand the problem of concurrent computing, it
is important to have a basic understanding of the HVM.

@@

Again, square brackets and parentheses work the same in Harmony, so
\texttt{[a, b, c]} (which looks like a Python list)
is the same Harmony value as \texttt{(a, b, c)} (which looks like a Python tuple),
which in turn is the same Harmony value as \texttt{dict\{ 0:a, 1:b, 2:c \}}.
So if $x$ = \texttt{[ False, True ]},
then $x$[0] = \texttt{False} and $x$[1] = \texttt{True}, just like in Python.
However, when creating a singleton list, make sure you include the
comma, as in \texttt{[False,]}.

Harmony is not an object-oriented language, so objects don't have
built-in methods.  However, Harmony does have some powerful operators to
make up for some of that.
For example, dictionaries have two handy unary operators.
If $d$ is a
dictionary, then \texttt{keys}~$d$ (or equivalently \texttt{keys}($d$)
returns the set of keys and \texttt{len}~$d$ returns the size of
this set.

@@

A Harmony program is translated
into \emph{bytecode}, which is list of machine instructions that the
Harmony virtual machine (HVM) executes.
The HVM is not an ordinary virtual machine, but its architecture
is nonetheless representative of conventional computers and
virtual machines such as the Java Virtual Machine.

@@

The state of a process is called a \emph{context} (aka \emph{continuation}):
\index{context}%
\index{continuation}%
it contains the value of
its name tag, program counter, stack, and register.
The state of a HVM
consists of the value of its memory and the multiset (or \emph{bag})
\index{bag}%
\index{multiset}%
of contexts.  It is a multiset of contexts because two different processes can
be in the same state.
The initial state of the Harmony memory is the empty dictionary, \texttt{dict\{\}}.
The context bag has an initial context in it with name tag
\texttt{\_\_main\_\_/()}, pc 0, register \texttt{()}, and an empty stack.
Each machine instruction updates the state in some way.
% \autoref{fig:incstate} shows an example of a reachable state for the program in \autoref{fig:inc}.

@@

It may seem strange that there is only one memory location and that each
process has only one register.  However, this is not a limitation because
Harmony values are unbounded trees.
Both the memory and the register of a process always contain
a dictionary that maps atoms to Harmony values.  We call this a \emph{directory}.
\index{directory}%
A directory represents the state of a collection of variables named by the atoms.
%
Because directories are Harmony values themselves,
directories can be organized into a tree.
Each node in a directory tree is then identified
by a sequence of atoms, like a path name in the file system hierarchy.  We call
such a sequence the \emph{address}
\index{address}%
of a Harmony value, and it is relative to the
``root'' of the directory tree.

@@

Compiling the code in \autoref{fig:inc} results in the HVM bytecode
listed in \autoref{fig:inccode}.
You can obtain this code by invoking \texttt{harmony} with the \texttt{-a} flag
like so:
\begin{code}
\begin{verbatim}
harmony -a Up.hny
\end{verbatim}
\end{code}
Each process in the HVM is predominantly a \emph{stack machine},
\index{stack machine}%
but it also has a register.
All instructions are atomically executed.
Most instructions pop values from the stack or push values onto the stack.
At first there is one process, with name tag \texttt{\_\_init\_\_/()},
that initializes the state.
It starts executing at instruction 0 and keeps executing until the last
execution in the program.
In this case, the first instruction is a \texttt{JUMP} instruction that sets the
program counter to 39.
At program counter 1 is the code for the \texttt{incrementer} method.
All methods start with a \texttt{Frame} instruction and end with a \texttt{Return}
instruction.

@@

\begin{enumerate} \setcounter{enumi}{1}
\item The \texttt{Load} instruction pushes the value of the
\textit{count} variable onto the stack.
\item the \texttt{Push} instruction pushes the constant 1
onto the stack of the process.
\item \texttt{2-ary} is a \texttt{+} operation with 2 arguments.
It pops two values from the stack (1 and the value of \textit{count}),
adds them and pushes the result back onto the stack.
\item The \texttt{Store} instruction pops
a Harmony value (the sum of the \textit{count} variable and 1) and
stores it in the \textit{count} variable.
\end{enumerate}

>>>>

Harmony programs are compiled to Harmony \emph{bytecode}
\index{bytecode}%
(a list of machine instructions for a virtual machine),
which in turn is executed by the Harmony virtual machine (HVM).
\index{virtual machine}%
\index{Harmony Virtual Machine}%
\index{HVM}%
To understand the problem of concurrent computing, it
is important to have a basic understanding of the HVM.

@@

Again, square brackets and parentheses work the same in Harmony, so
\texttt{[a, b, c]} (which looks like a Python list)
is the same Harmony value as \texttt{(a, b, c)} (which looks like a Python tuple),
which in turn is the same Harmony value as \texttt{dict\{ 0:a, 1:b, 2:c \}}.
So if $x$ = \texttt{[ False, True ]},
then $x$[0] = \texttt{False} and $x$[1] = \texttt{True}, just like in Python.
However, when creating a singleton list, make sure you include the
comma, as in \texttt{[False,]}.  The expression \texttt{[False]} just means
\texttt{False}.

Harmony is not an object-oriented language, so objects don't have
built-in methods.  However, Harmony does have some powerful operators to
make up for some of that.
For example, dictionaries have two handy unary operators.
If $d$ is a
dictionary, then \texttt{keys}~$d$ (or equivalently \texttt{keys}($d$))
returns the set of keys and \texttt{len}~$d$ returns the size of
this set.

@@

A Harmony program is translated into HVM bytecode.
The HVM is not an ordinary virtual machine, but its architecture
is nonetheless representative of conventional computers and
virtual machines such as the Java Virtual Machine.

@@

The state of a process is called a \emph{context} (aka \emph{continuation}):
\index{context}%
\index{continuation}%
it contains the value of
its name tag, program counter, stack, and register.
The state of a HVM
consists of the value of its memory and the multiset (or \emph{bag})
\index{bag}%
\index{multiset}%
of contexts.  It is a multiset of contexts because two different processes can
be in the same state.
The initial state of the Harmony memory is the empty dictionary, \texttt{dict\{\}}.
The context bag has an initial context in it with name tag
\texttt{\_\_init\_\_/()}, pc 0, register \texttt{()}, and an empty stack.
Each machine instruction updates the state in some way.
% \autoref{fig:incstate} shows an example of a reachable state for the program in \autoref{fig:inc}.

@@

It may seem strange that there is only one memory location and that each
process has only one register.  However, this is not a limitation because
Harmony values are unbounded trees.
Both the memory and the register of a process always contain
a dictionary that maps atoms to Harmony values.  We call this a \emph{directory}.
\index{directory}%
A directory represents the state of a collection of variables named by the atoms.
%
Because directories are Harmony values themselves,
directories can be organized into a tree.
Each node in a directory tree is then identified
by a sequence of atoms, like a path name in the file system hierarchy.  We call
such a sequence the \emph{address}
\index{address}%
of a Harmony value, and it is relative to the
``root'' of the directory tree.
For example, in \autoref{fig:inc} the memory is a dictionary with two
entries: \texttt{.count} and \texttt{.done}.  And the value of entry
\texttt{.done} is a dictionary with keys 0 and 1.

@@

Compiling the code in \autoref{fig:inc} results in the HVM bytecode
listed in \autoref{fig:inccode}.
You can obtain this code by invoking \texttt{harmony} with the \texttt{-a} flag
like so:
\begin{code}
\begin{verbatim}
harmony -a Up.hny
\end{verbatim}
\end{code}
Each process in the HVM is predominantly a \emph{stack machine},
\index{stack machine}%
but it also has a register.
All instructions are atomically executed.
Most instructions pop values from the stack or push values onto the stack.
At first there is one process, with name tag \texttt{\_\_init\_\_/()},
that initializes the state.
It starts executing at instruction 0 and keeps executing until the last
execution in the program.
In this case, the first instruction is a \texttt{JUMP} instruction that sets the
program counter to 34 (skipping over the code for the methods).
At program counter 1 is the code for the \texttt{incrementer} method.
All methods start with a \texttt{Frame} instruction and end with a \texttt{Return}
instruction.

@@

\begin{enumerate} \setcounter{enumi}{1}
\item The \texttt{Load} instruction pushes the value of the
\textit{count} variable onto the stack.
\item the \texttt{Push} instruction pushes the constant 1
onto the stack of the process.
\item \texttt{2-ary} is a \texttt{+} operation with 2 arguments.
It pops two values from the stack (the value of \textit{count} and 1),
adds them, and pushes the result back onto the stack.
\item The \texttt{Store} instruction pops
a Harmony value (the sum of the \textit{count} variable and 1) and
stores it in the \textit{count} variable.
\end{enumerate}
